taxize - taxonomic search and retrieval in R

### 2013-07-05 inst/doc/taxize_withcode_plos.Rnw
started work on matching example

\subsection*{Matching species tables with different taxonomic resolution}
Trait-based approaches are a promising tool in ecology. Unlike taxonomy-based methods, traits may not be constrained to biogeographic boundaries \cite{baird_toward_2011} and have potential to disentangle the effects of multiple stressors \cite{statzner_can_2010}.

To analyse trait-composition abundance data must be matched with trait databases like \cite{usseglio-polatera_biological_2000}. However these two data tables may contain species information on different taxonomic levels and perhabs data must be aggregated to a joint taxomic level. Therefore we have to search the taxomomic ladder up- and downwards for matches.

taxize can help in this data-cleaning step, providing a reproducible workflow. This is illustrated on a small fictitious example in the supplement.

\subsection*{Aggregating data to a specific taxonomic rank}

In biology, one can asks questions at varying taxonomic levels. One may perform analyses on different taxonomic levels. This use case is easily handled in taxize. A function called \emph{tax\_agg} will aggregate community data to a specific taxonomic level. In this example, we take data of 5 species and aggregate them to family level. Again we can specify if we want to use data from ITIS or NCBI.

>>>>

\subsection*{Matching species tables with different taxonomic resolution}
Biologist often need to match different sets of data tied to species. Trait-based approaches are a promising tool in ecology \cite{statzner_can_2010}. One problem is that abundance data must be matched with trait databases like \cite{usseglio-polatera_biological_2000}. These two data tables may contain species information on different taxonomic levels and perhabs data must be aggregated to a joint taxomic level, so that the data can be merged. taxize can help in this data-cleaning step, providing a reproducible workflow.

<<tax_match, cache=TRUE>>=
"<TODO>"
@

A more complicated example is given in Appendix B.

\subsection*{Aggregating data to a specific taxonomic rank}
In biology, one can asks questions at varying taxonomic levels. One may perform analyses on different taxonomic levels. This use case is easily handled in taxize. A function called \emph{tax\_agg} will aggregate community data to a specific taxonomic level. In this example, we take data of 5 species and aggregate them to family level. Again we can specify if we want to use data from ITIS or NCBI.

### 2013-07-05 inst/doc/taxize_withcode_plos.tex
started work on matching example

The R language is the dominant language used by biologists (reference), and now has over 5,000 packages on the R package repository (CRAN) and more than 2,500 packages on other repositories to extend R. R is great for manipulating, visualizing and fitting statistical models to data. However, the key missing piece in R is the ability to get data from the internet within R. Getting data from the web will be increasingly common as more and more data gets moved to the cloud. Increasingly, data is available from the web via API's, or application programming interfaces. These are bits of code that allow computers to talk to one another using code that is not human readable, but is machine readable. Web APIs often define a number of methods that allow users to search for a species name, or retrieve the synonyms for a species name, for example. A further strength of APIs is that they are language agnostic, meaning that data can be consumed in almost any computing context, allowing users to interact with the web API without having to know the details of the code. Whereas, if data are stored in an Excel file, for example, the file can only be opened in a few programs.

@@

A number of data sources in taxize provide the capability to retrieve higher taxonomic names, but we will highlight two of the more useful ones: ITIS and NCBI. The principle in both is the same - first you need to get a numeric identifier for the queried species, then retrieve the higher taxonomy with this identifier. First, we'll explore the user of ITIS, by searching for two species, \emph{Abies procera} and \emph{Pinus contorta}.

\begin{knitrout}
\definecolor{shadecolor}{rgb}{0.969, 0.969, 0.969}\color{fgcolor}\begin{kframe}
\begin{alltt}
specieslist <- \hlfunctioncall{c}(\hlstring{"Abies procera"}, \hlstring{"Pinus contorta"})
\hlfunctioncall{classification}(\hlfunctioncall{get_tsn}(specieslist, \hlstring{"sciname"}))
\end{alltt}
\end{kframe}
\end{knitrout}

It turns out both species are in the family Pinaceae. You can also get this type of information from the NCBI by doing \emph{classification(get\_uid(specieslist))}.

Instead of a full classification, you may only want a single name, say a family name for your species of interest. The function \emph{tax\_name} is built just for this purpose. And you can specify the data source you retrieve the taxonomic name from with the \emph{db} parameter, either ITIS or NCBI.

\begin{knitrout}
\definecolor{shadecolor}{rgb}{0.969, 0.969, 0.969}\color{fgcolor}\begin{kframe}
\begin{alltt}
\hlfunctioncall{tax_name}(query = \hlstring{"Helianthus annuus"}, get = \hlstring{"family"}, db = \hlstring{"itis"})
\hlfunctioncall{tax_name}(query = \hlstring{"Helianthus annuus"}, get = \hlstring{"family"}, db = \hlstring{"ncbi"})
\end{alltt}
\end{kframe}
\end{knitrout}

\subsection*{Interactive name selection}
In the previous section we used the function \emph{get\_tsn} to get a classification for two plant species. Below are a few examples. When you run these examples in R, you are presented with a command prompt asking for the row that contains the name you would like back; that output is not printed below for brevity. In this example, the search term has many matches. The function returns a data.frame of the matches, and asks for the user to input what row number to accept.

\begin{knitrout}
\definecolor{shadecolor}{rgb}{0.969, 0.969, 0.969}\color{fgcolor}\begin{kframe}
\begin{alltt}
\hlfunctioncall{get_tsn}(searchterm = \hlstring{"Elisabethae"}, searchtype = \hlstring{"sciname"})
\end{alltt}
\end{kframe}
\end{knitrout}

@@

\begin{knitrout}
\definecolor{shadecolor}{rgb}{0.969, 0.969, 0.969}\color{fgcolor}\begin{kframe}
\begin{alltt}
\hlfunctioncall{get_uid}(sciname = \hlfunctioncall{c}(\hlstring{"Chironomus riparius"}, \hlstring{"aaa vva"}))
\end{alltt}
\end{kframe}
\end{knitrout}

\subsection*{Retrieve a phylogeny}
Ecologists are increasingly taking a phylogenetic approach to ecology, applying phylogenies to topics such as the study of community structure \cite{webb2002phylogenies}, ecological networks \cite{rafferty2013phylogenetic}, functional trait ecology \cite{REF}. Yet, Many biologists are not adequately trained in reconstructing phylogenies. Fortunately, there are some sources for getting a phylogeny without having to know how to build one; one of these is for angiosperms, called Phylomatic \cite{webb2005}. We have created a workflow in taxize that accepts a species list, and taxize works behind the scenes to get higher taxonomic names, which are required by Phylomatic to get a phylogeny. Here is a short example, producing the tree in figure \ref{fig:phylomatic}.

@@

\subsection*{What taxa are the children of my taxon of interest?}
If you aren't a taxonomic specialist on a particular taxon you likely don't know what children taxa are within a family, or within a genus. This task becomes especially unwieldy when there are a large number of taxa downstream. You can of course go to a website like Wikispecies (\cite{wikispecies}) or Encyclopedia of Life (\cite{eol}) to get downstream names. However, taxize provides an easy way to programatically search for downstream taxa, both for the Catalogue of Life (CoL; \cite{col}) and the Integrated Taxonomic Information System (\cite{itis}). Here is a short example using the CoL in which we want to find all the species within the genus \emph{Apis} (honey bees).

\begin{knitrout}
\definecolor{shadecolor}{rgb}{0.969, 0.969, 0.969}\color{fgcolor}\begin{kframe}
\begin{alltt}
\hlfunctioncall{col_downstream}(name = \hlstring{"Apis"}, downto = \hlstring{"Species"})[[1]]
\end{alltt}
\end{kframe}
\end{knitrout}

@@

\subsection*{IUCN Status}
There are a number of things we can do once we have the correct taxonomic names. One thing we can do is ask about the conservation status of a species. We have provided a set of functions, \emph{iucn\_summary} and \emph{iucn\_status}, to search for species names, and extract the status information, respectively. Here, we search for the Panther and Lynx.

\begin{knitrout}
\definecolor{shadecolor}{rgb}{0.969, 0.969, 0.969}\color{fgcolor}\begin{kframe}
\begin{alltt}
ia <- \hlfunctioncall{iucn_summary}(\hlfunctioncall{c}(\hlstring{"Panthera uncia"}, \hlstring{"Lynx lynx"}))
\hlfunctioncall{iucn_status}(ia)
\end{alltt}
\end{kframe}
\end{knitrout}

@@

\begin{knitrout}
\definecolor{shadecolor}{rgb}{0.969, 0.969, 0.969}\color{fgcolor}\begin{kframe}
\begin{alltt}
out[\hlfunctioncall{grep}(\hlstring{"RAG1"}, out$genesavail, ignore.case = T), ]
\end{alltt}
\end{kframe}
\end{knitrout}

It turns out that there are XX different unique records found. However, this doesn't mean that there are XX different genes found as the API does not provide metadata to classify genes. However, at the end of the example, we showed that you can use regular expressions via (e.g., via \emph{grep}) to search for the gene of interest.

\subsection*{Matching species tables with different taxonomic resolution}
Trait-based approaches are a promising tool in ecology. Unlike taxonomy-based methods, traits may not be constrained to biogeographic boundaries \cite{baird_toward_2011} and have potential to disentangle the effects of multiple stressors \cite{statzner_can_2010}.

To analyse trait-composition abundance data must be matched with trait databases like \cite{usseglio-polatera_biological_2000}. However these two data tables may contain species information on different taxonomic levels and perhabs data must be aggregated to a joint taxomic level. Therefore we have to search the taxomomic ladder up- and downwards for matches.

taxize can help in this data-cleaning step, providing a reproducible workflow. This is illustrated on a small fictitious example in the supplement.

>>>>

The R language is the dominant language used by biologists (reference), and now has over 4,500 packages on the Comprehensive R Archive Network (CRAN) and more than 2,500 packages on other repositories to extend R. R is great for manipulating, visualizing and fitting statistical models to data. However, the key missing piece in R is the ability to get data from the internet within R. Getting data from the web will be increasingly common as more and more data gets moved to the cloud. Increasingly, data is available from the web via application programming interfaces (API). These are bits of code that allow computers to talk to one another using code that is not human readable, but is machine readable. Web APIs often define a number of methods that allow users to search for a species name, or retrieve the synonyms for a species name, for example. A further strength of APIs is that they are language agnostic, meaning that data can be consumed in almost any computing context, allowing users to interact with the web API without having to know the details of the code. Whereas, if data are stored in an Excel file, for example, the file can only be opened in a few programs.

@@

A number of data sources in taxize provide the capability to retrieve higher taxonomic names, but we will highlight two of the more useful ones: Integrated Taxonomic Information System (ITIS) \cite{itis} and National Center for Biotechnology Information (NCBI) \cite{federhen}. First, we'll search for two species, \emph{Abies procera} and \emph{Pinus contorta}.

\begin{knitrout}
\definecolor{shadecolor}{rgb}{0.969, 0.969, 0.969}\color{fgcolor}\begin{kframe}
\begin{alltt}
specieslist <- \hlfunctioncall{c}(\hlstring{"Abies procera"}, \hlstring{"Pinus contorta"})
\hlfunctioncall{classification}(specieslist, db = \hlstring{"itis"})
\end{alltt}
\begin{verbatim}
[[1]]
        rankName       taxonName    tsn
1        Kingdom         Plantae 202422
2     Subkingdom  Viridaeplantae 846492
3   Infrakingdom    Streptophyta 846494
4       Division    Tracheophyta 846496
5    Subdivision Spermatophytina 846504
6  Infradivision    Gymnospermae 846506
7          Class       Pinopsida 500009
8          Order         Pinales 500028
9         Family        Pinaceae  18030
10         Genus           Abies  18031
11       Species   Abies procera 181835

[[2]]
        rankName       taxonName    tsn
1        Kingdom         Plantae 202422
2     Subkingdom  Viridaeplantae 846492
3   Infrakingdom    Streptophyta 846494
4       Division    Tracheophyta 846496
5    Subdivision Spermatophytina 846504
6  Infradivision    Gymnospermae 846506
7          Class       Pinopsida 500009
8          Order         Pinales 500028
9         Family        Pinaceae  18030
10         Genus           Pinus  18035
11       Species  Pinus contorta 183327
\end{verbatim}
\end{kframe}
\end{knitrout}

It turns out both species are in the family Pinaceae. You can also get this type of information from the NCBI by doing \emph{classification(specieslist, db = 'ncbi')}.

Instead of a full classification, you may only want a single name, say a family name for your species of interest. The function \emph{tax\_name} is built just for this purpose. As with the \emph{classification}-function you can specify the data source with the \emph{db} argument, either ITIS or NCBI.

\begin{knitrout}
\definecolor{shadecolor}{rgb}{0.969, 0.969, 0.969}\color{fgcolor}\begin{kframe}
\begin{alltt}
\hlfunctioncall{tax_name}(query = \hlstring{"Helianthus annuus"}, get = \hlstring{"family"}, db = \hlstring{"itis"})
\end{alltt}
\begin{verbatim}
      family
1 Asteraceae
\end{verbatim}
\begin{alltt}
\hlfunctioncall{tax_name}(query = \hlstring{"Helianthus annuus"}, get = \hlstring{"family"}, db = \hlstring{"ncbi"})
\end{alltt}
\begin{verbatim}
      family
1 Asteraceae
\end{verbatim}
\end{kframe}
\end{knitrout}

\subsection*{Interactive name selection}
As mentioned most databases use a numeric code to reference a species. A general workflow in taxize is: Retrieve Code for the queried species and then use this code to query more data/information.
Below are a few examples. When you run these examples in R, you are presented with a command prompt asking for the row that contains the name you would like back; that output is not printed below for brevity. In this example, the search term has many matches. The function returns a data.frame of the matches, and asks for the user to input what row number to accept.

\begin{knitrout}
\definecolor{shadecolor}{rgb}{0.969, 0.969, 0.969}\color{fgcolor}\begin{kframe}
\begin{alltt}
\hlfunctioncall{get_tsn}(searchterm = \hlstring{"Heliastes"}, searchtype = \hlstring{"sciname"})
\end{alltt}
\begin{verbatim}
           combinedname    tsn
1     Heliastes bicolor 615238
2   Heliastes chrysurus 615250
3     Heliastes cinctus 615573
4  Heliastes dimidiatus 615257
5  Heliastes hypsilepis 615273
6 Heliastes immaculatus 615639
7 Heliastes opercularis 615300
8      Heliastes ovalis 615301
 1 
NA 
attr(,"class")
[1] "tsn"
\end{verbatim}
\end{kframe}
\end{knitrout}

@@

\begin{knitrout}
\definecolor{shadecolor}{rgb}{0.969, 0.969, 0.969}\color{fgcolor}\begin{kframe}
\begin{alltt}
\hlfunctioncall{get_uid}(sciname = \hlfunctioncall{c}(\hlstring{"Chironomus riparius"}, \hlstring{"aaa vva"}))
\end{alltt}
\begin{verbatim}
[1] "315576" NA      
attr(,"class")
[1] "uid"
\end{verbatim}
\end{kframe}
\end{knitrout}

\subsection*{Retrieve a phylogeny}
Ecologists are increasingly taking a phylogenetic approach to ecology, applying phylogenies to topics such as the study of community structure \cite{webb2002phylogenies}, ecological networks \cite{rafferty2013phylogenetic}, functional trait ecology \cite{poff2006functional}. Yet, Many biologists are not adequately trained in reconstructing phylogenies. Fortunately, there are some sources for getting a phylogeny without having to know how to build one; one of these is for angiosperms, called Phylomatic \cite{webb2005}. We have created a workflow in taxize that accepts a species list, and taxize works behind the scenes to get higher taxonomic names, which are required by Phylomatic to get a phylogeny. Here is a short example, producing the tree in figure \ref{fig:phylomatic}.

\begin{knitrout}
\definecolor{shadecolor}{rgb}{0.969, 0.969, 0.969}\color{fgcolor}\begin{kframe}
\begin{alltt}
taxa <- \hlfunctioncall{c}(\hlstring{"Poa annua"}, \hlstring{"Abies procera"}, \hlstring{"Helianthus annuus"})
tree <- \hlfunctioncall{phylomatic_tree}(taxa = taxa)
tree$tip.label <- \hlfunctioncall{capwords}(tree$tip.label)
\hlfunctioncall{plot}(tree, cex = 1)
\end{alltt}
\end{kframe}
\end{knitrout}

@@

\subsection*{What taxa are the children of my taxon of interest?}
If someone is not a taxonomic specialist on a particular taxon he likely does not know what children taxa are within a family, or within a genus. This task becomes especially unwieldy when there are a large number of taxa downstream. You can of course go to a website like Wikispecies \cite{wikispecies} or Encyclopedia of Life \cite{eol} to get downstream names. However, taxize provides an easy way to programatically search for downstream taxa, both for the Catalogue of Life (CoL) \cite{col} and the Integrated Taxonomic Information System \cite{itis}. Here is a short example using the CoL in which we want to find all the species within the genus \emph{Apis} (honey bees).

\begin{knitrout}
\definecolor{shadecolor}{rgb}{0.969, 0.969, 0.969}\color{fgcolor}\begin{kframe}
\begin{alltt}
\hlfunctioncall{col_downstream}(name = \hlstring{"Apis"}, downto = \hlstring{"Species"})[[1]]
\end{alltt}
\begin{verbatim}
  childtaxa_id     childtaxa_name childtaxa_rank
1      6971712 Apis andreniformis        Species
2      6971713        Apis cerana        Species
3      6971714       Apis dorsata        Species
4      6971715        Apis florea        Species
5      6971716 Apis koschevnikovi        Species
6      6845885     Apis mellifera        Species
7      6971717   Apis nigrocincta        Species
\end{verbatim}
\end{kframe}
\end{knitrout}

@@

\subsection*{IUCN Status}
There are a number of things we can do once we have the correct taxonomic names. One thing we can do is ask about the conservation status of a species (IUCN Red List of Threatened Species \cite{iucn}). We have provided a set of functions, \emph{iucn\_summary} and \emph{iucn\_status}, to search for species names, and extract the status information, respectively. Here, we search for the Panther and Lynx.

\begin{knitrout}
\definecolor{shadecolor}{rgb}{0.969, 0.969, 0.969}\color{fgcolor}\begin{kframe}
\begin{alltt}
ia <- \hlfunctioncall{iucn_summary}(\hlfunctioncall{c}(\hlstring{"Panthera uncia"}, \hlstring{"Lynx lynx"}))
\hlfunctioncall{iucn_status}(ia)
\end{alltt}
\begin{verbatim}
Panthera uncia      Lynx lynx 
          "EN"           "LC" 
\end{verbatim}
\end{kframe}
\end{knitrout}

@@

\begin{knitrout}
\definecolor{shadecolor}{rgb}{0.969, 0.969, 0.969}\color{fgcolor}\begin{kframe}
\begin{alltt}
out[\hlfunctioncall{grep}(\hlstring{"RAG1"}, out$genesavail, ignore.case = TRUE), ]
\end{alltt}
\begin{verbatim}
        spused length
413 Umbra limi    732
427 Umbra limi    959
434 Umbra limi   1631
                                                                            genesavail
413 isolate UlimA recombinase activating protein 1 (rag1) gene, exon 3 and partial cds
427           recombination-activating protein 1 (RAG1) gene, intron 2 and partial cds
434                        recombination-activating protein 1 (RAG1) gene, partial cds
    access_num       ids
413   JX190826 394772608
427   AY459526  45479841
434   AY380548  38858304
\end{verbatim}
\end{kframe}
\end{knitrout}

It turns out that there are XX different unique records found. However, this doesn't mean that there are XX different genes found as the API does not provide metadata to classify genes. However, at the end of the example, we showed that you can use regular expressions (e.g., via \emph{grep}) to search for the gene of interest.

\subsection*{Matching species tables with different taxonomic resolution}
Biologist often need to match different sets of data tied to species. Trait-based approaches are a promising tool in ecology \cite{statzner_can_2010}. One problem is that abundance data must be matched with trait databases like \cite{usseglio-polatera_biological_2000}. These two data tables may contain species information on different taxonomic levels and perhabs data must be aggregated to a joint taxomic level, so that the data can be merged. taxize can help in this data-cleaning step, providing a reproducible workflow.

\begin{knitrout}
\definecolor{shadecolor}{rgb}{0.969, 0.969, 0.969}\color{fgcolor}\begin{kframe}
\begin{alltt}
\hlstring{"<TODO>"}
\end{alltt}
\begin{verbatim}
[1] "<TODO>"
\end{verbatim}
\end{kframe}
\end{knitrout}

A more complicated example is given in Appendix B.

\subsection*{Aggregating data to a specific taxonomic rank}
In biology, one can asks questions at varying taxonomic levels. One may perform analyses on different taxonomic levels. This use case is easily handled in taxize. A function called \emph{tax\_agg} will aggregate community data to a specific taxonomic level. In this example, we take data of 5 species and aggregate them to family level. Again we can specify if we want to use data from ITIS or NCBI.

\begin{knitrout}
\definecolor{shadecolor}{rgb}{0.969, 0.969, 0.969}\color{fgcolor}\begin{kframe}
\begin{alltt}
\hlfunctioncall{data}(dune, package = \hlstring{'vegan'})
df <- dune[ , 1:5]
\hlfunctioncall{colnames}(df) <- \hlfunctioncall{c}(\hlstring{"Bellis perennis"}, \hlstring{"Empetrum nigrum"}, \hlstring{"Juncus bufonius"}, 
                  \hlstring{"Juncus articulatus"}, \hlstring{"xxx"})
\hlfunctioncall{head}(df)
\end{alltt}
\begin{verbatim}
   Bellis perennis Empetrum nigrum Juncus bufonius Juncus articulatus xxx
2                3               0               0                  0   0
13               0               0               3                  0   0
4                2               0               0                  0   0
16               0               0               0                  3   0
6                0               0               0                  0   0
1                0               0               0                  0   0
\end{verbatim}
\begin{alltt}
agg <- \hlfunctioncall{tax_agg}(df, rank = \hlstring{'family'}, db = \hlstring{'ncbi'})
agg
\end{alltt}
\begin{verbatim}

	Aggregated community data

Level of Aggregation: FAMILY
No. taxa before aggregation: 5
No. taxa after aggregation: 4
No. taxa not found: 1
\end{verbatim}
\begin{alltt}
\hlfunctioncall{head}(agg$x)
\end{alltt}
\begin{verbatim}
   Asteraceae Ericaceae Juncaceae xxx
2           3         0         0   0
13          0         0         3   0
4           2         0         0   0
16          0         0         3   0
6           0         0         0   0
1           0         0         0   0
\end{verbatim}
\end{kframe}
\end{knitrout}

We see that the two Juncus species are aggregated to Juncaceae and their abundances are summed up. If a taxon is one lower taxonomic resolution then the querried or the taxon is not found in the database (like in the example taxon \emph{xxx}) then these are not aggregated and returned as is.
